---
title: "Project memo"
author: "Group 7"
output: github_document
---

This document should contain a detailed account of the data clean up for your data and the design choices you are making for your plots. For instance you will want to document choices you've made that were intentional for your graphic, e.g. color you've chosen for the plot. Think of this document as a code script someone can follow to reproduce the data cleaning steps and graphics in your handout.

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(readxl)
library(usmap)
library(ggrepel)
```

## Data Clean Up Steps for Overall Data

```{r general_data_cleaning}
setwd("/cloud/project")

energy <- read_csv("data/energy.csv")
filtered_energy <- energy %>%
  filter(Year == 2019) %>%
  select(State, Year, starts_with(c("Consumption", "Expenditure", "Price")))

population <- read_excel("data/Population.xlsx")
filtered_population <- population[c(-(1:8),-(60:66)),] %>%
  rename(Population = "...13",
         State = "table with row headers in column A and column headers in rows 3 through 4. (leading dots indicate sub-parts)") %>%
  mutate(State = substr(State, 2, nchar(State))) %>%
  select(State, Population)

joined <- filtered_energy %>%
  left_join(filtered_population, by = c("State" = "State"))

variables <- colnames(joined)
variables <- variables[-c(1:2,84)]

per_cap <- joined %>%
  mutate(across(variables, ~ .x / joined$Population))

for (var in variables) {
  per_cap[[paste0(var, "_pc")]] <- per_cap[[var]]
}

per_cap <- per_cap %>%
  select(State, Year, ends_with("pc"))

final <- joined %>%
  left_join(per_cap, by = c("State" = "State"))

# Print the output of glimpse() or skim()
glimpse(final)
```

## Plots

### ggsave example for saving plots

```{r starwars ggsave example}
p1 <- starwars |>
  filter(mass < 1000, 
         species %in% c("Human", "Cerean", "Pau'an", "Droid", "Gungan")) |>
  ggplot() +
  geom_point(aes(x = mass, 
                 y = height, 
                 color = species)) +
  labs(x = "Weight (kg)", 
       y = "Height (m)",
       color = "Species",
       title = "Weight and Height of Select Starwars Species",
       caption = paste("This data comes from the starwars api: https://swapi.py43.com"))


ggsave("example-starwars.png", width = 4, height = 4)

ggsave("example-starwars-wide.png", width = 6, height = 4)
```



### Plot 1: Commercial Petroleum Consumption Map

#### Data cleanup specific to plot 1

```{r adjust_state_label}
final <- final %>%
  mutate(state = State)
```


#### Final Plot 1:

```{r creat_map_petroleum, warning=FALSE}
plot_usmap(
  linewidth = 0.1,
  color = "white",
  regions = "state",
  data = final,
  values = 'Consumption.Commercial.Petroleum_pc'
) +
  scale_fill_gradient(
    labels = scales::label_number(big.mark = ','),
    high = 'steelblue',
    low = 'white'
  ) +
  labs(fill = 'Consumption (Billions BTU)',
       title = "Commercial Petroleum Consumption by State Across U.S.A.") +
  guides(
    fill = guide_colorbar(barwidth = unit(8, 'cm'))
  ) +
  theme(
    legend.position = 'top',
    plot.title = element_text(hjust = 0.5)  # <-- Center the title
  )
```


### Plot 2: Commercial Petroleum Consumption Per Capita Map 

```{r creat_map_petroleum_pc, warning=FALSE}
plot_usmap(
  linewidth = 0.1,
  color = "white",
  regions = "state",
  data = final,
  values = 'Expenditure.Commercial.Petroleum_pc'
) +
  scale_fill_gradient(
    labels = scales::label_number(big.mark = ','),
    high = 'steelblue',
    low = 'white'
  ) +
  labs(fill = 'Consumption (Billions BTU)',
       title = "Commercial Petroleum Consumption by State Across U.S.A.") +
  guides(
    fill = guide_colorbar(barwidth = unit(8, 'cm'))
  ) +
  theme(
    legend.position = 'top',
    plot.title = element_text(hjust = 0.5)  # <-- Center the title
  )
```

### Plot 3: Petroleum Expenditure Per Capita vs Petroleum Dependance

#### Data cleanup specific to plot 3

#Note: rowSums, grepl, contains, and ends_with are functions for creating aggreates of multiple variables based on naming conventions learned from ECON 368 with Professor Kyle Coombs.

```{r create_expenditure_plot_data}

expenditure_plot_data <- final %>%
  mutate(
    total_consumption_pc = rowSums(select(., ends_with("_pc")), na.rm = TRUE),
    
    petroleum_consumption_pc = rowSums(
      select(., contains("Consumption") & contains("Petroleum") & ends_with("_pc")), na.rm = TRUE),
    
    petroleum_dependence = petroleum_consumption_pc / total_consumption_pc,
    
    petroleum_expenditure_pc = rowSums(
      select(., contains("Expenditure") & contains("Petroleum") & ends_with("_pc")), na.rm = TRUE)
  )

```

#### Final Plot 3:

```{r expenditure_plot}

expenditure_plot <- ggplot(expenditure_plot_data, aes(x = petroleum_dependence,
                        y = petroleum_expenditure_pc,
                        label = State)) +
  geom_point(size = 3, alpha = 0.8, color = "steelblue") +   # steel blue
  geom_smooth(method = "lm", color = "blue", linewidth = 1, se = TRUE) +
  geom_text_repel(size = 3, color = "black") +
  labs(
    title = "Petroleum Expenditure per Capita vs. Petroleum Dependence",
    x = "Proportion of Total State Consumption of Petroleum",
    y = "Petroleum Expenditure per Capita (USD)"
  ) +
  theme_minimal(base_size = 11.5)

expenditure_plot

```


### Plot 4: Petroleum Price vs. Petroleum Dependance

#### Data cleanup specific to plot 3

```{r create_price_plot_data}

price_plot_data <- final %>%
  mutate(
    total_consumption_pc = rowSums(select(., ends_with("_pc")), na.rm = TRUE),
    
    petroleum_consumption_pc = rowSums(
      select(., contains("Consumption") & contains("Petroleum") & ends_with("_pc")), na.rm = TRUE),
    
    petroleum_dependence = petroleum_consumption_pc / total_consumption_pc,
    
    petroleum_price = rowMeans(
      select(., contains("Price") & contains("Petroleum")), na.rm = TRUE)
  )

```

#### Final Plot 1:

```{r price_plot}

price_plot <- ggplot(price_plot_data, aes(x = petroleum_dependence,
                        y = petroleum_price,
                        label = State)) +
  geom_point(size = 3, alpha = 0.8, color = "steelblue") +
  geom_smooth(method = "lm", color = "blue", linewidth = 1, se = TRUE) +
  geom_text_repel(size = 3, color = "black") +
  labs(
    x = "Proportion of Total State Consumption from Petroleum",
    y = "Average Petroleum Price (USD per million BTU)"
  ) +
  theme_minimal(base_size = 11.5) +
  theme(
    axis.text = element_text(size = 10.5),
    axis.title = element_text(size = 10.5)
  )

price_plot

```


